\large Neural and Symbolic Optimization of Cosmological Particle-Mesh Simulation \vspace{0.3em}

### 2023-09-06 docs/papers/sto/so.tex
FIXUP updating draft

\subsection{Optimization of Dynamics}
In cosmological $N$-body simulations, the dynamics of the particles is governed
by the Poisson equation
%
\begin{equation}
  \nabla^2 \varphi(\vx) = \frac32 \Omegam H_0^2 \delta(\vx),
\end{equation}
%
In Fourier space, the gravitational force field $- \vnabla \varphi$ reads
%
\begin{equation}
- i \vk \varphi(\vk) = i \frac32 \Omegam H_0^2 \,\frac{\vk}{k^2} \delta(\vk),
\end{equation}
%
We introduce force sharpening through
%
\begin{equation}
\frac{i k_i}{k^2} \to \frac{i k_i}{k^2}
  f(k_i; \vvartheta) g(k_1, k_2, k_3; \vvartheta),
\end{equation}
%
where $f$ and $g$ are NNs, each approximating a nonlinear function of the
wavevector components and all possible relevant features $\vvartheta$ discussed
in \autoref{sec:features}.
Being embedded in the differentiable simulation, the NNs can be trained using
gradient-based optimization methods as in usual deep learning tasks.
The parameters of the neural networks, which enter the force evaluation in
every time integration step, can be treated just the same as other cosmological
parameters.
The objective gradients are accumulated, based on which the parameters are then
optimized.
For both $f$ and $g$ neural networks, we adopt multilayer perceptron (MLP, i.e.
fully connected) structures.

We generate our target data for training and testing with the \GADGET\
simulation code \citep{GADGET-4}.

@@

\vspace{1em}
\subsection{Differentiable Simulation}

The differentiable cosmological simulation with adjoint method \citep{Li2022a}
and the corresponding PM $N$-body library \pmwd\ \citep{Li2022b} are the
foundation of the neural optimization discussed in this work.
Within this fully differentiable framework, we are able to embed neural networks
in any physical process and train them for many purposes, including the
optimization of the gravitational force field in this work.

@@

\begin{table}
\centering
\caption{TODO Possible features that can affect force sharpening.
All scales are then multiplied by the wavenumber $k$ to form dimensionless
factors.}
\label{tab:feat}
\begin{tabular}{rl}
\toprule
type & feature \\
\midrule
& $R_\mathrm{d}$: RMS linear displacement \\
\cmidrule(lr){2-2}
& $R_P$: $\Delta_\linear^2(R_P^{-1}) \triangleq 1$ \\
\cmidrule(lr){2-2}
nonlinear scales & $R_\tophat$: $\sigma_\tophat^2(R_\tophat) \triangleq 1$ \\
\cmidrule(lr){2-2}
& $R_\gauss$: $\sigma_\gauss^2(R_\gauss) \triangleq 1$ \\
\cmidrule(lr){2-2}
& $\d R / \d\lna$: time derivatives of the above \\
\midrule
& particle spacing \\
\cmidrule(lr){2-2}
other scales & $l$: PM force mesh cell size \\
\cmidrule(lr){2-2}
& softening length \\
\midrule
& $G_m(a) \triangleq D_m(a) / a^m, m \in \{1, 2, \YL{3's?}\}$ \\
\cmidrule(lr){2-2}
& $\d\ln G_m / \d\lna$ \\
\cmidrule(lr){2-2}
dimensionless factors & $\Omegam(a)$ \\
\cmidrule(lr){2-2}
& $\d\ln\!H / \d\lna$ \\
\cmidrule(lr){2-2}
& $\Delta\lna$: time step size in $\lna$ \\
\bottomrule
\end{tabular}
\end{table}

Relevant typical scales are defined through
\begin{itemize}
\item The RMS linear displacement is given by
%
\begin{equation}
R_\mathrm{d} = \int \frac{k P_\linear}{6\pi^2} \d\lnk.
\end{equation}
%
\item The dimensionless power spectrum is defined as
%
\begin{equation}
\Delta^2(k) \triangleq \frac{k^3 P(k)}{2 \pi^2}.
\end{equation}
%
\item The variance of linear matter overdensity in a top-hat window is given by
%
\begin{equation}
\sigma_\tophat^2(R) = \int_0^\infty \frac{\d k}k
  \Delta_\linear^2(k) W_\tophat^2(kR),
\end{equation}
%
where $W_\tophat(kR) = 3[\sin(kR) - kR\cos(kR)] / (kR)^3$ is the
Fourier transform of the top-hat window.
\item Likewise $\sigma_\gauss^2(R)$ is defined with a Gaussian window
$W_\gauss(kR) = e^{-(kR)^2/2}$.
\end{itemize}

@@

We combine the above list of parameters in $\vvartheta$.

Dimensionless combinations by multiplying the wavevector components with the
scales, e.g.,
%
\begin{equation}
f(k_i; \vvartheta) = f(k_iR_d, k_iR_\tophat, \cdots; G_m, \Omegam, \cdots)
\end{equation}
%
\begin{equation}
g(k_1, k_2, k_3; \vvartheta) = g(\{k_jR_d, k_jR_\tophat, \cdots\}_{j=1,2,3}; G_m, \Omegam, \cdots)
\end{equation}
%
$f$ and $g$ are optimized during training to correct and sharpen the
gravitational forces.
Gravity obeys the $\mathrm{O}(3)$ rotational symmetry, which is however
broken by the PM solver to the O\textsubscript{h} discrete symmetry with
48 elements involving permutations and reversals of the 3 principal axes
of the simulation box.
The PM evaluation of each force component can further break the symmetry
along that axis, leaving the remaining 2 permutationally symmetric.
Therefore, we implement $f$ and $g$ to be even functions of the
wavevector components, and $g$ to be permutation equivariant in all 3
components, so that $f g$ is the most general modification to the force
kernel.
An even and real kernel in configuration space is also even in Fourier
space.
We take the absolute value for the evenness, and sort the 3 wavevector
compontents to achieve the permutation equivariance.

@@

We run 512 \GADGET\ simulations that cover a wide range of
configurations and cosmological parameters, as listed in
\autoref{tab:param}.
Each simulation has $128^3$ particles and writes 121 snapshots linearly
spaced in scale factor from $1/16 + \Delta a$ to $1 + \Delta a$, with
spacing $1/128$.
$\Delta a$ is an offset, common for all snapshots in each simulation,
that varies as in \autoref{tab:param} and helps to sample different
alignment between snapshot and start/stop times.

>>>>

\subsection{Optimization of Dynamics}

\begin{table}
  \centering
  \caption{Possible features that can affect force sharpening.
  All scales are then multiplied by the wavenumber $k$ to form dimensionless
  factors.}
  \label{tab:feat}
  \begin{tabular}{rl}
  \toprule
  type & feature \\
  \midrule
  & $R_\mathrm{d}$: RMS linear displacement \\
  \cmidrule(lr){2-2}
  & $R_P$: $\Delta_\linear^2(R_P^{-1}) \triangleq 1$ \\
  \cmidrule(lr){2-2}
  nonlinear scales & $R_\tophat$: $\sigma_\tophat^2(R_\tophat) \triangleq 1$ \\
  \cmidrule(lr){2-2}
  & $R_\gauss$: $\sigma_\gauss^2(R_\gauss) \triangleq 1$ \\
  \cmidrule(lr){2-2}
  & $\d R / \d\lna$: time derivatives of the above \\
  \midrule
  & particle spacing \\
  \cmidrule(lr){2-2}
  other scales & $l$: PM force mesh cell size \\
  \cmidrule(lr){2-2}
  & softening length \\
  \midrule
  & $G_m(a) \triangleq D_m(a) / a^m, m \in \{1, 2, \YL{3's?}\}$ \\
  \cmidrule(lr){2-2}
  & $\d\ln G_m / \d\lna$ \\
  \cmidrule(lr){2-2}
  dimensionless factors & $\Omegam(a)$ \\
  \cmidrule(lr){2-2}
  & $\d\ln\!H / \d\lna$ \\
  \cmidrule(lr){2-2}
  & $\Delta\lna$: time step size in $\lna$ \\
  \bottomrule
  \end{tabular}
  \end{table}

In cosmological $N$-body simulations, the dynamics of the particles is governed
by the Poisson equation
%
\begin{equation}
  \nabla^2 \varphi(\vx) = \frac32 \Omegam H_0^2 \delta(\vx),
\end{equation}
%
In Fourier space, the gravitational force field $- \vnabla \varphi$ reads
%
\begin{equation}
- i \vk \varphi(\vk) = i \frac32 \Omegam H_0^2 \,\frac{\vk}{k^2} \delta(\vk),
\end{equation}
%
We introduce force sharpening through
%
\begin{equation}
\frac{i k_i}{k^2} \to \frac{i k_i}{k^2}
  f(k_i; \vvartheta) g(k_1, k_2, k_3; \vvartheta),
\end{equation}
%
where $f$ and $g$ are NNs, each approximating a nonlinear function of the
wavevector components and all possible relevant features $\vvartheta$ summarized
in \autoref{tab:feat} and more details discussed in \autoref{sec:features}.
Gravity obeys the $\mathrm{O}(3)$ rotational symmetry, which is however
broken by the PM solver to the O\textsubscript{h} discrete symmetry with
48 elements involving permutations and reversals of the 3 principal axes
of the simulation box.
The PM evaluation of each force component can further break the symmetry
along that axis, leaving the remaining 2 permutationally symmetric.
Therefore, we implement $f$ and $g$ to be even functions of the
wavevector components, and $g$ to be permutation equivariant in all 3
components, so that $f g$ is the most general modification to the force
kernel.
An even and real kernel in configuration space is also even in Fourier
space.
We take the absolute value for the evenness, and sort the 3 wavevector
compontents to achieve the permutation equivariance.
Being embedded in the differentiable simulation, the NNs can be trained using
gradient-based optimization methods as in usual deep learning tasks.
The parameters of the neural networks, which enter the force evaluation in
every time integration step, can be treated just the same as other cosmological
parameters.
The objective gradients are accumulated, based on which the parameters are then
optimized.
For both $f$ and $g$ neural networks, we adopt multilayer perceptron (MLP, i.e.
fully connected) structures.

We generate our target data for training and testing with the \GADGET\
simulation code \citep{GADGET-4}.
We run 512 \GADGET\ simulations that cover a wide range of
configurations and cosmological parameters, as listed in
\autoref{tab:param}.
Each simulation has $128^3$ particles and writes 121 snapshots linearly
spaced in scale factor from $1/16 + \Delta a$ to $1 + \Delta a$, with
spacing $1/128$.
$\Delta a$ is an offset, common for all snapshots in each simulation,
that varies as in \autoref{tab:param} and helps to sample different
alignment between snapshot and start/stop times.
In our \pmwd\ model, we generate the snapshot at an arbitrary given time using
the cubic Hermite interpolation, with more details in \autoref{sec:snapobs}.

@@

\vspace{1em}
\subsection{Differentiable Simulation}
\label{sec:pmwd}

The differentiable cosmological simulation with adjoint method \citep{Li2022a}
and the corresponding PM $N$-body library \pmwd\ \citep{Li2022b} are the
foundation of the neural optimization discussed in this work.
Within this fully differentiable framework, we are able to embed neural networks
in any physical process and train them for many purposes, including the
optimization of the gravitational force evaluation in this work.

@@

Multiple scales and dimensionless factors could possibly affect force
sharpening, which are denoted as $\vvartheta$ and enter the NNs of $f(k_i;
\vvartheta)$ and $g(k_1, k_2, k_3; \vvartheta)$ as input features.
A list of these features is summarized in \autoref{tab:feat}, and here we
present relevant definations and more details.

Below are some definations related to the typical scales.
\begin{itemize}
\item The root-mean-square (RMS) linear displacement is given by
%
\begin{equation}
R_\mathrm{d} = \int \frac{k P_\linear(k)}{6\pi^2} \d\lnk,
\end{equation}
%
where $P_\linear(k)$ is the linear power spectrum.
\item The dimensionless linear power spectrum is defined as
%
\begin{equation}
\Delta^2_\linear(k) \triangleq \frac{k^3 P_\linear(k)}{2 \pi^2},
\end{equation}
%
based on which we define a nonlinear scale $R_P$ through
$\Delta_\linear^2(R_P^{-1}) \triangleq 1$.
\item The variance of linear matter overdensity in a top-hat window is given by
%
\begin{equation}
\sigma_\tophat^2(R) = \int_0^\infty \frac{\d k}k
  \Delta_\linear^2(k) W_\tophat^2(kR),
\end{equation}
%
where $W_\tophat(kR) = 3[\sin(kR) - kR\cos(kR)] / (kR)^3$ is the
Fourier transform of the top-hat window.
Then we have $R_\tophat$: $\sigma_\tophat^2(R_\tophat) \triangleq 1$.
\item Likewise $\sigma_\gauss^2(R)$ is defined with a Gaussian window
$W_\gauss(kR) = e^{-(kR)^2/2}$, and $R_\gauss$: $\sigma_\gauss^2(R_\gauss)
\triangleq 1$.
\end{itemize}
These nonlinear scales are functions of time since the linear power spectrum
evolves with time through the growth factor $D(a)$, we also take the time
derivatives of these scales as input features.
Other relevant scales include the particle spacing in Lagrangian space, the cell
size of the chosen mesh grid for force evaluation, and the softening length in
the target \GADGET\ simulations.

@@

Dimensionless combinations are formed by multiplying the wavevector components
with the scales, i.e.,
%
\begin{equation}
f(k_i; \vvartheta) = f(k_iR_d, k_iR_P, \cdots; G_m, \Omegam, \cdots),
\end{equation}
%
\begin{equation}
g(k_1, k_2, k_3; \vvartheta) = g(\{k_jR_d, k_jR_P, \cdots\}_{j=1,2,3}; G_m, \Omegam, \cdots).
\end{equation}
%

@@
